---
title: "Gentle introduction to social networks in R"
output: html_notebook
---

### Prof. Jose Manuel MAGALLANES. 
* Professor at Pontificia Universidad Católica del Peru 
* Senior Data Scientist at eScience Institute, University of Washington

_______


## CONTENTS:

### 1. Network Creation from dataframes

These data frames have been previously prepared, based on the data from [50 influential people in Seattle tech to follow on Twitter](https://www.builtinseattle.com/2018/08/06/50-seattle-tech-twitter-accounts-to-follow). The file can be seen [here](https://docs.google.com/spreadsheets/d/e/2PACX-1vTpWvtfphnO9eRYwmEyxVAxxo1KEsdpK6sK6q5uhJn5x2QtB-eGiso8ibpF16NaHAers4wDIHkvBo64/pubhtml), and includes 3 sheets:

* Network as two columns representing edges.
* Network as adjacency matrix.
* Attributes of the Network nodes.


```{r}
rm(list = ls())
##
fileLink='https://github.com/eScience-UW/gentleIntro_networks/raw/main/data/seattleTop.xlsx'

edges=rio::import(fileLink, which = 'edges') 
adjacency=rio::import(fileLink, which = 'adjacency')
attributes=rio::import(fileLink,which ="attributes")
```

Use the previous dataframes to create the network:

* From the edges:

```{r}
library(igraph) # package needed

EliteNet=graph_from_data_frame(edges, directed = F)

summary(EliteNet)
```

* From adjacency matrix:

```{r}
#prepare matrix
row.names(adjacency)=adjacency[,1] #first column for index
adjacency=adjacency[,-1] # deleting the first column
adjacencyMatrix=as.matrix(adjacency) 

# the making...
EliteNet=graph_from_adjacency_matrix(adjacencyMatrix,mode="undirected")


summary(EliteNet)
```
See inside:

```{r}
V(EliteNet)
```

```{r}
E(EliteNet)
```

### 2. Adding Attributes

Attributes are added using **vectors**:

```{r}
# edges
EliteNet$weight=1
# EliteNet=set_edge_attr(EliteNet,"weight", value = 1)

#nodes
V(EliteNet)$male=attributes$male
V(EliteNet)$followers=attributes$followers
# EliteNet=set_vertex_attr(EliteNet, "followers", value=attributes$followers)
```

You can use attributes to color nodes:

```{r}
library(ggraph)

base=ggraph(graph = EliteNet)
base + geom_node_label(aes(label = name,
                           color=as.factor(male)),
                       repel = TRUE,show.legend = F) + 
       geom_edge_link(alpha=0.1) + 
       scale_color_manual(values = c('red','blue'))
  
```


## 3. Exploring the Network


* Connectedness: A network is  “connected” if there exists a _path_ between any pair of nodes (undirected networks).

```{r}
is_connected(EliteNet)
```
* Density: from 0 to 1, where 1 makes it a ‘complete’ network: there is a link between every pair of nodes.

```{r}
graph.density(EliteNet)
```

* Diameter: When two vertices are connected, one can reach the other using multiple egdes. The geodesic is the shorthest path between two connected vertices. Then, the diameter, is the maximum geodesic in a network.

```{r}
diameter(EliteNet)
```

* Assortativity: it is a measure to see if nodes are connecting to other nodes similar to themselves. Closer to 1 means higher assortativity, closer to -1 diassortativity; while 0 is no assortativity.

    - Degree assortativity: tendency for highly connected nodes to be connected among themselves.

```{r}
assortativity_degree(EliteNet)
```
    - Categorical assortativity: tendency for nodes with other nodes sharing the same category.
    
```{r}
assortativity_nominal(EliteNet,as.factor(V(EliteNet)$male),directed = F)
```
    - Assortativity (numerical): 

```{r}
assortativity(EliteNet,V(EliteNet)$followers,directed = T)
```
## 4. Exploration of Network nodes

- The eigenvector of a vertex will tell you how well connected is a vertex; that is, vertices with the highest values are considered the most influential as they are connected to vertices that are also well connected.

- The closeness of a vertex will tell you how close is a vertex to every other vertex. A vertex with high closeness can share information faster than the rest.

- The betweeness of a vertex will tell you how critical is a vertex to connect vertex that are not connected directly.


```{r}

eigen=eigen_centrality (EliteNet,scale=F)$vector

close=closeness(EliteNet,normalized=T)

betw=betweenness(EliteNet,normalized=T)

DFCentrality=as.data.frame(cbind(eigen,close,betw),stringsAsFactors = F)
names(DFCentrality)=c('Eigenvector','Closeness','Betweenness')

DFCentrality$person=row.names(DFCentrality)
row.names(DFCentrality)=NULL

```

```{r}
library(ggplot2)
ggplot(DFCentrality, aes(x=Betweenness, y=Closeness)) + 
    theme_classic() +
    geom_text(aes(label=person,size=Eigenvector),show.legend = F,alpha=0.5) 
```
The previous results tells us that two people are salient:

```{r}
HubNodes=dplyr::top_n(DFCentrality,2,Eigenvector)$person
HubNodes
```
Let's highlight those guys:

```{r, fig.height=10}
NodeCount=length(V(EliteNet))

V(EliteNet)$label=''

for (index in seq(1:NodeCount)){
    currentName=V(EliteNet)$name[index]
    if (currentName%in%HubNodes){
            V(EliteNet)$label[index]=currentName
    }
}

```


```{r}
library(ggraph)

base=ggraph(graph = EliteNet)
base  + geom_node_label(aes(label = label), 
                        repel = TRUE,
                        show.legend = F,
                        color='red') + 
         geom_edge_link(alpha=0.1)
```


## 5. Communities

```{r}
RandomNet <- erdos.renyi.game(length(V(EliteNet)), 0.5)
```

* Transitivity: How probable is that two nodes with a common connection, are also connected.

```{r}
transitivity(RandomNet)
```
```{r}
transitivity(EliteNet)
```
* Partition

Set of nodes that belong to a group. The hard problems is to decide what node goes where:

```{r}
# edges that will partition the net
mincut=igraph::min_cut(EliteNet, value.only = FALSE)$cut
mincut
```
```{r}
labelsBye=unique(as.vector(ends(EliteNet,mincut))) # flattening
labelsBye
```
```{r}
EliteNet_cut=delete_edges(EliteNet,mincut)

V(EliteNet)$label=''

for (index in seq(1:NodeCount)){
    currentName=V(EliteNet)$name[index]
    if (currentName%in%labelsBye){
            V(EliteNet)$label[index]=currentName
    }
}

base=ggraph(graph =EliteNet_cut ) + geom_edge_link(alpha=0.2)
base + geom_node_label(aes(label=label),repel = T) 
```

The **girvan_newman** algorithm is based on creating partitions using the centrality of nodes to create partition too. We can create something similar to the last result like this:


```{r}
# creating all partitions
partition_girvanNewman_all <- edge.betweenness.community(EliteNet,merges = T)

# the first partition
partition_girvanNewman_first= cut_at(partition_girvanNewman_all,2)
```

```{r}
ColorIndex=partition_girvanNewman_first
```

```{r}
base=ggraph(graph =EliteNet) + geom_edge_link(alpha=0.2)

base + geom_node_point(aes(color=ColorIndex), show.legend = F, size=4) +
    geom_node_text(aes(label=label,color=ColorIndex),repel = T,show.legend = F) 
```


A more complex algorithm is the [Louvain](https://perso.uclouvain.be/vincent.blondel/research/louvain.html): 

```{r}
partition_louvain=cluster_louvain(EliteNet)
```

```{r}
V(EliteNet)$louvain=membership(partition_louvain)

base=ggraph(graph =EliteNet) + geom_edge_link(alpha=0.2)
base + geom_node_point(aes(color=as.factor(louvain)), show.legend = F, size=4) 

```

```{r}
a_partition_girvanNewman=cut_at(partition_girvanNewman_all,max(membership(partition_louvain)))
```

* Modularity:

If we get positive values (being 1 the top value), we could consider there are good community structures ([wiki](https://en.wikipedia.org/wiki/Modularity_(networks))). The higher modularity the connections between the nodes are dense  within a partition but sparse  between nodes in different partitions.

```{r}
modularity(EliteNet, a_partition_girvanNewman)
```

```{r}
modularity(EliteNet,membership(partition_louvain))
```

```{r}
V(EliteNet)$louvain=membership(partition_louvain)
V(EliteNet)$girvanNewman=membership(partition_louvain)
```


```{r, eval=TRUE}
write_graph(EliteNet, "EliteNet_R.graphml", "graphml")
```
